---
title: "Descriptive analysis for macronutrient PGS x SES analysis"
output: html_document
---

```{r, include=FALSE, echo=F, error=F, warning=F, message=F}

knitr::opts_chunk$set(echo=F, error=F, message=F, warning=F, warning=F, fig.align = "center",
                      out.width = "90%", out.height = "50%", fig.height=5)

## load packages & pantry scripts
library(tidyverse) 
library(data.table)
library(ggpubr)
library(knitr)
library(ComplexHeatmap)

lapply(list.files("../../pantry/functions/", full.names = T), source)


## Preference phenos/PGS vars & labels 
pheno_labs <- c(pref_sweet_food="Prefer sweet food", pref_bitter_food="Prefer bitter food", 
               pref_salty_food="Prefer salty food", pref_fatty_food="Prefer fatty food")
phenoPalettes.l <- list(
  pref_sweet_food=c(palettes$NatExt$Reds[c(2,5)]), pref_bitter_food=c(palettes$NatExt$Blues[c(2,5)]),
  pref_salty_food=c(palettes$NatExt$Oranges[c(2,5)]), pref_fatty_food=c(palettes$NatExt$Purples[c(2,5)]) )

pgs_labs <- paste0(as.vector(pheno_labs), " PGS")
names(pgs_labs) <- paste0(names(pheno_labs),"_pgs")
pgsPalettes <- phenoPalettes.l ; names(pgsPalettes)<-paste0(names(phenoPalettes.l),"_pgs")

pgs_palettes <- c(pref_sweet_pgs=palettes$NatExt$Reds[3], pref_bitter_pgs=palettes$NatExt$Blues[3],
                  pref_salty_pgs=palettes$NatExt$Oranges[3], pref_fatty_food=palettes$NatExt$Purples[3])
pheno_palettes <- pgs_palettes ; names(pheno_palettes) <- gsub("pgs","food",names(pgs_palettes))

pgs <- names(pgs_labs)
pgs_short=gsub(" food","",pgs_labs)
prefs <- names(pheno_labs)
prefs_short <- gsub(" food","",pheno_labs)
  
tastes <- c("bitter","sweet","salt","sour","umami", "fatty")

plot_theme <- theme_bw() + theme(
  axis.text = element_text(size=6), 
  axis.title = element_text(size=8)
  ) 

```

```{r}
analysis <- readRDS("../data/processed/ukb_MA_analysis_testing.rda") 
analysis <- analysis %>% 
  drop_na(paste0(pgs,"_decile")) %>%
  mutate(glu2hr=ifelse(fasting_hrs<=2,glu,NA),
         glufast5hr=ifelse(fasting_hrs>=5,glu,NA))
  
#fullsample <- readRDS("../data/processed/ukb_MA_analysis_allphenos.rda")
```


### UKB participant characteristics

```{r}
left_join(
  fread("../data/results/descr_tab_basic_testing.csv", header=T),
  fread("../data/results/descr_tab_basic_training.csv", header=T), by="V1"
) %>% fwrite("../data/results/descr_tab_basic_traintest_compare.csv")

```

```{r}
fread("../data/results/descr_tab_allPGS_descr_byQuintile_testing.csv", header=T) %>% select(-Total) %>% kable()
fread("../data/results/descr_tab_allPGS_diet24hr_byQuintile_testing.csv", header=T) %>% select(-Total) %>% kable()
fread("../data/results/descr_tab_allPGS_diet_byQuintile_testing.csv", header=T) %>% select(-Total) %>% kable()
fread("../data/results/descr_tab_allPGS_cmrf_byQuintile_testing.csv", header=T) %>% select(-Total) %>% kable()
```


## Histograms of preference phenotypes & PGS
##### Phenotype
```{r, fig.height=4}
# preference phenotypes
plot_pref_hists.l <- lapply(1:4, function(p) {
  plot_continuous(prefs[p], data=analysis, fill=phenoPalettes[[p]][1]) + ggtitle(" ") + ylim(0,8500) +
    xlab(pheno_labs[p]) + plot_theme
}) ; ggarrange(plotlist = plot_pref_hists.l, nrow=1) #%>% 
  #ggsave(filename="../output/plot_macro_pheno_histograms_testing.pdf", height=3.5, width=8)
```
##### Minimally-adjusted
```{r}
pgs_adjMin <- paste0(pgs, "_adjMin")
plot_pgs_hists.l <- lapply(1:4, function(p) {
  plot_continuous(pgs_adjMin[p], data=analysis, fill=pgsPalettes[[p]][1]) + ggtitle(" ") + ylim(0,4500) +
    xlab(pgs_adjMin[p]) + plot_theme
}) ; ggarrange(plotlist = plot_pgs_hists.l, nrow=1) %>% 
  ggsave(filename="../output/plot_macro_pgs_adjMin_histograms_testing.pdf", height=3, width=8)

```
##### Behaviorally-adjusted
```{r}
plot_pref_boxp <- 
  analysis %>% select(id, starts_with("pref") & ends_with("food")) %>%
    filter(complete.cases(.)) %>%
    pivot_longer(-id) %>%
    mutate(name=factor(name, levels=names(pheno_labs), labels=pheno_labs)) %>%
    ggplot(aes(x=name, y=value, fill=name)) + 
    geom_boxplot(alpha=0.8) + 
    scale_fill_manual(values=as.vector(pheno_palettes), name="Preference Trait") + 
    scale_y_continuous(limits=c(0,9), breaks=seq(0,9,1)) + 
    theme_bw() + 
    ylab("Preference (0-9)") + xlab("") + 
  plot_theme + theme(axis.text = element_text(size=8)) + 
  theme(axis.title = element_text(size=9, face="bold", vjust=2))

plot_pref_boxp %>% ggsave(filename = "../output/plot_macro_pheno_boxplots_testing.pdf", height=3.5, width=8)

# preference pgs
plot_pgs_hists.l <- lapply(1:4, function(p) {
  plot_continuous(pgs[p], data=analysis, fill=pgsPalettes[[p]][1]) + ggtitle(" ") + ylim(0,4000) +
    xlab(pgs_labs[p]) + plot_theme
}) ; ggarrange(plotlist = plot_pgs_hists.l, nrow=1) %>% 
  ggsave(filename="../output/plot_macro_pheno_histograms_testing.pdf", height=3, width=8)

```

## PGS distributions by race/ethnicity

```{r}
#Ethnicity
p_pgs_ethn <- analysis %>% select(ethnicity, all_of(pgs)) %>%
  pivot_longer(-ethnicity) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=ethnicity, y=value, color=PGS)) + 
  geom_boxplot() + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  xlab("Ethicity") + ylab("Food Preference PGS") +
  theme(axis.text.x = element_text(angle=90, size=8, hjust=1, vjust=1)) ; p_pgs_ethn

p_pgs_anc <- analysis %>% select(ancestry, all_of(pgs)) %>%
  pivot_longer(-ancestry) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=ancestry, y=value, color=PGS)) + 
  geom_boxplot() + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  xlab("Genetic Ancestry") + ylab("Food Preference PGS") +
  theme(axis.text.x = element_text(angle=90, size=8, hjust=1, vjust=1)) ; p_pgs_anc

p_pgs_ethn %>% ggsave(filename="../output/plot_pgsPref_ethnicity_v1.pdf", height=4, width=7)
p_pgs_anc %>% ggsave(filename="../output/plot_pgsPref_ancestry_v1.pdf", height=3, width=7)
```


##### By Education & Income Level
```{r}
p_pgs_educ_4lvl <- analysis %>% select(educ_4level.lab, all_of(pgs)) %>%
  pivot_longer(-educ_4level.lab) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=educ_4level.lab, y=value, color=PGS)) + 
  geom_boxplot() + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  xlab("Education Level (4-Level)") + ylab("Food Preference PGS") +
  theme(axis.text.x = element_text(angle=90, size=8, hjust=1, vjust=1)) ; p_pgs_educ_4lvl

p_pgs_educ_4lvl %>% ggsave(filename="../output/plot_pgsPref_educ_v1.pdf", height=3, width=7)

p_pgs_inc <- analysis %>% select(income_level.lab, all_of(pgs)) %>%
  pivot_longer(-income_level.lab) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=income_level.lab, y=value, color=PGS)) + 
  geom_boxplot() + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  xlab("Income Level") + ylab("Food Preference PGS") +
  theme(axis.text.x = element_text(angle=90, size=8, hjust=1, vjust=1)) ; p_pgs_inc

p_pgs_inc %>% ggsave(filename="../output/plot_pgsPref_inc_v1.pdf", height=3, width=7)

```

```{r}
#dot plot of associations between PGS with age & BMI

p_pgs_age <- analysis %>% select(pgs, age) %>%
  pivot_longer(-age) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=value, y=age, color=PGS)) + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  geom_point(size=0.75, alpha=0.75) +
  xlab("Preference PGS") + ylab("Age, years") +
  theme(axis.text.x = element_text(size=8)) ; p_pgs_age

p_pgs_age %>% ggsave(filename="../output/plot_pgsPref_age_v1.pdf", height=3, width=8)

p_pgs_bmi <- analysis %>% select(pgs, bmi) %>%
  pivot_longer(-bmi) %>%
  mutate(PGS=factor(name, levels=pgs, labels=pgs_labs)) %>%
  ggplot(aes(x=value, y=bmi, color=PGS)) + 
  facet_wrap(~PGS, nrow=1) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  ggtheme_basic_bw + 
  geom_point(size=0.75, alpha=0.75) +
  xlab("Preference PGS") + ylab("BMI, kg/m2") +
  theme(axis.text.x = element_text(size=8)) ; p_pgs_bmi

p_pgs_bmi %>% ggsave(filename="../output/plot_pgsPref_bmi_v1.pdf", height=3, width=8)

```

## Correlations of PGS with dietary traits
##### Heat Map for Behaviorally-adjusted
```{r, fig.height=8, fig.width=6, include=F}
cor_pgs_diet<-fread("../data/results/descr_cormat_pgs_allDiet_testing.csv")

# HeatMap with 24HR Food Groups
cor_pgs_24hr <- cor_pgs_diet %>% select(starts_with("fg") & -starts_with("pref") & -ends_with("pgs")) %>% as.matrix()
cor_pgs_ffq <- cor_pgs_diet %>% select(-starts_with("fg") & -starts_with("pref") & -ends_with("pgs")) %>% as.matrix()
cor_pgs_prefs <- cor_pgs_diet %>% select(-starts_with("fg") & starts_with("pref") & -ends_with("pgs")) %>% as.matrix()

ComplexHeatmap::Heatmap(t(cor_pgs_24hr), row_names_gp = gpar(fontsize = 5),
                        column_labels = pgs_short,
                        column_names_gp = gpar(fontsize=8), column_names_rot = T, 
                        column_names_centered = T) 

```

##### Validation of Behaviorally-adjusted PGS
```{r, simple_corr_scatter, echo=F}

prefs_palette <- as.vector(do.call(rbind,pgsPalettes)[,1])
names(prefs_palette) <- pheno_labs

plot_prefXpgs <- function(pref_pgs) {
  analysis %>% 
    rename(PGS=paste0(pref_pgs,"_decile")) %>%
    group_by(PGS=as.factor(PGS)) %>%
    reframe(sweet_mean=mean(pref_sweet_food, na.rm=T), sweet_se=sd(pref_sweet_food, na.rm=T),
            bitter_mean=mean(pref_bitter_food, na.rm=T), bitter_se=sd(pref_bitter_food, na.rm=T),
            salty_mean=mean(pref_salty_food, na.rm=T), salty_se=sd(pref_salty_food, na.rm=T),
            fatty_mean=mean(pref_fatty_food, na.rm=T), fatty_se=sd(pref_fatty_food, na.rm=T)) %>%
    pivot_longer(-"PGS", names_sep="_", names_to=c("prefer", "msr")) %>%
    pivot_wider(names_from=msr, values_from = value) %>%
    mutate(Prefs=factor(paste("Prefer", prefer, "food"), levels=pheno_labs)) %>%
    ggplot(aes(x=PGS, y=mean, color=Prefs, ymin=mean-se, ymax=mean+se)) +
    geom_line(aes(group=prefer), linewidth=0.25) +
    geom_point(size=1.25, alpha=0.9) + 
    theme_bw() + plot_theme + ylim(3,7) +
    xlab(pgs_labs[pref_pgs])+ ylab("Mean Food Preference Score (0-9)") + 
    scale_color_manual(values=prefs_palette, guide=NULL) + 
    theme(panel.grid.major.x = element_blank())
}

ggpubr::ggarrange(plotlist=lapply(pgs, plot_prefXpgs), nrow=1, common.legend = T, legend = "right") %>%
  ggsave(filename="../output/plot_descr_pgs_foodpref_testing.pdf", height=3, width=8)
```

##### Validation of minimallyy-adjusted PGS
```{r, simple_corr_scatter, echo=F}
names(pgs_adjMin)<-pgs_labs
ggpubr::ggarrange(plotlist=lapply(pgs_adjMin, plot_prefXpgs), nrow=1, common.legend = T, legend = "right") 

```


## R2 (variability explained) in macronutrient intake by PGS

##### Behaviorally-adjusted
```{r}
demo=paste0("age+sex+",paste0("PC",1:10,collapse = "+"))
bmi=paste0(demo, "+bmi")
life=paste0(bmi, "+smoke_level.lab+alch_level.lab+sleep+health")
ses=paste0(life, "+educ_level.lab+townsend")
covarsets.l <- list(Demographic=demo, BMI=bmi, Lifestyle=life, SES=ses)

comp_pheno_pgs <- do.call(rbind.data.frame, lapply(1:4, function(p) {
  y=prefs[[p]] ; Y=names(pheno_labs)[p] ; x=pgs[[p]] ; X=names(pgs_labs)[p]
  baseModel <- summary(lm(formula(paste0(y, "~", x )), data=analysis))
  do.call(rbind.data.frame, lapply(1:length(covarsets.l), function(P) {
    covModel <- summary(lm(formula(paste0(y, "~", covarsets.l[[P]] )), data=analysis))
    adjModel <- summary(lm(formula(paste0(y, "~", x, "+", covarsets.l[[P]] )), data=analysis))
    rbind.data.frame(
      cbind.data.frame(pref=Y, pgs=X, model=names(covarsets.l)[P], r2=covModel$r.squared),
      cbind.data.frame(pref=Y, pgs=X, model=paste0(names(covarsets.l)[P], "+PGS"), r2=adjModel$r.squared) )
  })) %>% rbind.data.frame(cbind.data.frame(pref=Y, pgs=X, model="PGS", r2=baseModel$r.squared))
}))
  
p_comp_r2 <- comp_pheno_pgs %>%
  mutate(pgs=factor(pgs, levels=names(pgs_labs), labels=pgs_labs)) %>%
  mutate(model = factor(model, levels=c("PGS", paste0(rep(names(covarsets.l),each=2),c("","+PGS"))))) %>% 
  mutate(groups = factor(c(rep(c(rep(names(covarsets.l),each=2),"PGS"),4)), levels=c("PGS", names(covarsets.l)))) %>%
  mutate(shapes = c(rep(c("A","B"),4),"B", rep(c("C","D"),4),"D", rep(c("E","F"),4),"F", rep(c("G","H"),4),"H")) %>%
  arrange(model) %>%
  ggplot(aes(x=groups, group=shapes, y=r2*100, fill=shapes)) + 
  geom_hline(yintercept = 0, color="black", linewidth=0.35) +
  facet_wrap(~pgs, scales="free_y",nrow=1) +
  geom_bar(stat="identity", position=position_dodge2(c(0.35, 0.25)),width=0.75) + 
  theme_bw() + scale_y_continuous(expand=c(0,0), limits=c(0,9.5), breaks=seq(0,9,1.5)) +
  scale_fill_manual(values=paste0(rep(as.vector(do.call(rbind,pgsPalettes)[,1]),each=2),rep(c("75", ""))),guide=NULL) + 
  scale_x_discrete(labels=c("PGS", "Basic", "BMI", "Lifestyle", "SES")) +
  ylab("R2 for Preference Trait") + xlab("Covariate Adjustments") +
  theme(axis.text.x = element_text(color="black", size=7),
        axis.title = element_text(size=8),
        axis.text.y = element_text(size=7),
        strip.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth=0.25))  #, angle=45, hjust =1, vjust=1)) + 
p_comp_r2 # %>% ggsave(filename="../output/plot_compare_r2_pgsVpheno_v1_freey_testing.pdf", height=3.25, width=10)
```

##### Behaviorally-adjusted
```{r}
comp_pheno_pgs_adjMin <- do.call(rbind.data.frame, lapply(1:4, function(p) {
  y=prefs[[p]] ; Y=names(pheno_labs)[p] ; x=pgs_adjMin[[p]] ; X=names(pgs_adjMin)[p]
  baseModel <- summary(lm(formula(paste0(y, "~", x )), data=analysis))
  do.call(rbind.data.frame, lapply(1:length(covarsets.l), function(P) {
    covModel <- summary(lm(formula(paste0(y, "~", covarsets.l[[P]] )), data=analysis))
    adjModel <- summary(lm(formula(paste0(y, "~", x, "+", covarsets.l[[P]] )), data=analysis))
    rbind.data.frame(
      cbind.data.frame(pref=Y, pgs=X, model=names(covarsets.l)[P], r2=covModel$r.squared),
      cbind.data.frame(pref=Y, pgs=X, model=paste0(names(covarsets.l)[P], "+PGS"), r2=adjModel$r.squared) )
  })) %>% rbind.data.frame(cbind.data.frame(pref=Y, pgs=X, model="PGS", r2=baseModel$r.squared))
}))
  
p_comp_r2_adjMin <- comp_pheno_pgs_adjMin %>%
  mutate(pgs=factor(pgs, levels=names(pgs_adjMin))) %>%
  mutate(model = factor(model, levels=c("PGS", paste0(rep(names(covarsets.l),each=2),c("","+PGS"))))) %>% 
  mutate(groups = factor(c(rep(c(rep(names(covarsets.l),each=2),"PGS"),4)), levels=c("PGS", names(covarsets.l)))) %>%
  mutate(shapes = c(rep(c("A","B"),4),"B", rep(c("C","D"),4),"D", rep(c("E","F"),4),"F", rep(c("G","H"),4),"H")) %>%
  arrange(model) %>%
  ggplot(aes(x=groups, group=shapes, y=r2*100, fill=shapes)) + 
  geom_hline(yintercept = 0, color="black", linewidth=0.35) +
  facet_wrap(~pgs, scales="free_y",nrow=1) +
  geom_bar(stat="identity", position=position_dodge2(c(0.35, 0.25)),width=0.75) + 
  theme_bw() + scale_y_continuous(expand=c(0,0), limits=c(0,9.5), breaks=seq(0,9,1.5)) +
  scale_fill_manual(values=paste0(rep(as.vector(do.call(rbind,pgsPalettes)[,1]),each=2),rep(c("75", ""))),guide=NULL) + 
  scale_x_discrete(labels=c("PGS", "Basic", "BMI", "Lifestyle", "SES")) +
  ylab("R2 for Preference Trait") + xlab("Covariate Adjustments") +
  theme(axis.text.x = element_text(color="black", size=7),
        axis.title = element_text(size=8),
        axis.text.y = element_text(size=7),
        strip.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth=0.25))  #, angle=45, hjust =1, vjust=1)) + 
p_comp_r2_adjMin # %>% ggsave(filename="../output/plot_compare_r2_pgsVpheno_v1_freey_testing.pdf", height=3.25, width=10)

ggarrange(p_comp_r2_adjMin,p_comp_r2, nrow=2, labels = c("Minimal Adjustment", "Behavioral Adjustment"),
          label.x = c(-0.07,-0.08), label.y=c(1.01,1.01), font.label = list(size=8)) %>%
  ggsave(filename="../output/plot_compare_r2_pgsVpheno_adjMin&adjBehav_20250413.pdf", height=5, width=10)
```


## Comparison of preference PGS with food preference traits

##### Behavioral-adjsuted 
```{r, fig.width=15, fig.height=6}
pgs_prefs <- fread("../data/results/lm_prefpgs_foodprefs_mBase_testing.csv")
codebook_prefs <- readxl::read_xlsx("../../tasteprefs/data/raw/food_pref_vars.xlsx")

prefs_taste <- codebook_prefs %>% pivot_longer(c(tastes, "nondiet", "spicy")) %>%
  filter(value == 1)

pgs_prefs_toPlot <- pgs_prefs %>% mutate(PGS=factor(exposure, levels=names(pgs_labs), labels=pgs_labs)) %>%
  mutate(dir=ifelse(beta>0,"pos","neg")) %>%
  rowwise() %>%
  mutate(cat = ifelse(outcome %in% prefs_taste$varID, prefs_taste$name[which(prefs_taste$varID==outcome)], "misc")) %>%
  ungroup() %>%
  mutate(cat=factor(str_to_sentence(cat), levels=c("Sweet", "Fatty", "Umami", "Bitter", "Salt", "Sour", "Spicy", "Misc", "Nondiet"))) %>%
  mutate(food=str_to_sentence(gsub("Liking for ", "", model)))

p <- pgs_prefs_toPlot %>%
  ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05/420), color="red") +
  geom_point(size=1.25, alpha=0.85) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=as.vector(pgs_palettes)) + 
  scale_color_manual(values=as.vector(pgs_palettes))

p %>% ggsave(filename="../output/plot_mht_prefPGS_foodprefs_testing.pdf", height=3, width=16)
```

## Comparison of preference PGS with FFQ dietary traits

##### Behavioral-adjusted & FFQ data
```{r, fig.width=15, fig.height=6}
pgs_ffq <- fread("../data/results/lm_prefpgs_ffqCont_mBase_testing.csv")
pgs_ffq_toPlot <- pgs_ffq %>% mutate(PGS=factor(exposure, levels=names(pgs_labs), labels=pgs_labs)) %>%
  mutate(dir=ifelse(beta>0,"pos","neg"))

p <- pgs_ffq_toPlot %>%
  mutate(food=factor(model, levels=diet_vars_cont)) %>%
  ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  #facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05), color="black", linetype="longdash") +
  geom_hline(yintercept = -log10(.05/(18*4)), color="red") +
  geom_point(size=1.55, alpha=0.85) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=as.vector(pgs_palettes)) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
  theme(panel.grid.minor =  element_blank(),
        panel.grid.major.x = element_blank()) + 
  coord_flip()

p %>% ggsave(filename="../output/plot_mht_prefPGS_ffq_testing.pdf", width=6, height=5)
```

##### Behavioral-adjusted & 24HR ata
```{r, fig.width=15, fig.height=6}
pgs_24hr <- fread("../data/results/lm_prefpgs_24hr_mBase_testing.csv")
pgs_24hr_toPlot <- pgs_24hr %>% mutate(PGS=factor(exposure, levels=names(pgs_labs), labels=pgs_labs)) %>%
  mutate(dir=ifelse(beta>0,"pos","neg"))

p <- pgs_24hr_toPlot %>%
  mutate(food=factor(model, levels=diet24hr_vars)) %>%
  ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  #facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05), color="black", linetype="longdash") +
  geom_hline(yintercept = -log10(.05/(15)), color="red") +
  geom_point(size=1.55, alpha=0.85) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=as.vector(pgs_palettes)) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
   theme(panel.grid.minor =  element_blank(),
        panel.grid.major.x = element_blank()) + 
  coord_flip()

p %>% ggsave(filename="../output/plot_mht_prefPGS_24hr_testing.pdf", width=7.5, height=5)
```


## Comparison of PGS with Macronutrient intake traits
```{r}
plot_intakeXpgs10 <- function(pref_pgs) {
  analysis %>% 
    rename(PGS=paste0(pref_pgs,"_decile")) %>%
    group_by(PGS=as.factor(PGS)) %>%
    reframe(Carb_mean=mean(carb_pheno, na.rm=T), Carb_se=sd(carb_pheno, na.rm=T),
            Fat_mean=mean(fat_pheno, na.rm=T), Fat_se=sd(fat_pheno, na.rm=T),
            Protein_mean=mean(prot_pheno, na.rm=T), Protein_se=sd(prot_pheno, na.rm=T)) %>%
    pivot_longer(-"PGS", names_sep="_", names_to=c("Macronutrient", "msr")) %>%
    pivot_wider(names_from=msr, values_from = value) %>%
    filter(!is.na(PGS)) %>%
    ggplot(aes(x=PGS, y=mean, color=Macronutrient, ymin=mean-se, ymax=mean+se)) +
    geom_point(size=2, position=position_dodge(0.35)) + geom_errorbar(width=0.25, position=position_dodge(0.35))+
    theme_bw() + plot_theme + 
    xlab(pgs_labs[pref_pgs]) + ylab("Macronutrient intake (%kcal), mean (SD) ") + 
    scale_color_manual(values=c("#A6501E", "#1D4884","#356932"))
}

ggpubr::ggarrange(plotlist=lapply(names(pgs_labs), plot_intakeXpgs10), nrow=1, common.legend = T, legend = "right") %>%
  ggsave(filename="../data/results/descr_plot_compare_pgs_macropct_testing.pdf", height=3, width=9)
```




## Preference PGS & CMRFs
Means/SE
```{r}
analysis_not2d <- analysis %>% filter(t2d_case.f=="Control")
pgs_deciles <- paste0(pgs,"_decile") ; pgs_quintiles <- paste0(pgs,"_quintile")

cmrf_pgs_forplot <- do.call(rbind.data.frame, lapply(pgs_deciles, function(pgs_dec) {
  do.call(rbind.data.frame, lapply(1:length(cmrf_vars), function(d) {
    analysis_not2d %>% rename(dec=pgs_dec) %>% group_by(PGS_quantile=all_of(dec)) %>%
      select(PGS_quantile, cmrf=names(cmrf_vars)[d]) %>%
      reframe(n=n(), mean=mean(cmrf, na.rm=T), sd=sd(cmrf, na.rm=T)) %>%
      mutate(se=sd/sqrt(n), 
             CMRF=cmrf_vars[d])
  })) %>% mutate(score=pgs_short[gsub("_decile","",pgs_dec)])
})) 

pgs_palette_d1d10 <- c(pgsPalettes$pref_sweet_food_pgs, pgsPalettes$pref_bitter_food_pgs, pgsPalettes$pref_salty_food_pgs)
p<-cmrf_pgs_forplot %>% 
  filter(PGS_quantile %in% c(1,10)) %>%
  filter(score != "Prefer fatty PGS") %>%
  mutate(Colors=factor(paste0(score, "-Q",PGS_quantile), levels=paste0(rep(pgs_short[1:3],each=2),c("-Q1","-Q10")))) %>%
  filter(CMRF %in% c( as.vector(c(cmrf_vars[c("hba1c","crp","hdl")])) ) ) %>%
  mutate(CMRF = factor(CMRF, levels=c("HbA1c, %", "CRP, mg/dL", "HDL, mmol/L", "BMI"))) %>%
  ggplot(aes(x=score, group=PGS_quantile, y=mean, ymin=mean-se, ymax=mean+se, color=Colors)) +
  facet_wrap(~CMRF, scale="free_y",nrow=1) +
  xlab("") + ylab("Mean (SE)") +
  geom_point(size=1, position = position_dodge(0.55)) + 
  geom_errorbar(width=0.45, position = position_dodge(0.55)) +
  scale_y_continuous(expand=c(0.15,0.01)) +
  scale_color_manual(values = c(rev(pgsPalettes$pref_sweet_food_pgs), 
                                rev(pgsPalettes$pref_bitter_food_pgs), 
                                rev(pgsPalettes$pref_salty_food_pgs))) + 
  theme_bw() +
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.x = element_text(angle=45,hjust=1,vjust=1,color="black",size=6),
        axis.text.y = element_text(size=5),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size=6),
        axis.title = element_text(size=6, color="black"),
        panel.spacing = unit(1.5, "lines"),
        legend.position = "none",
        plot.margin = margin(0.5,2.5,0.5,2.5,unit="line"))

  cmrf_pgs_forplot %>% filter(PGS_decile %in% c(1,10)) %>% filter(CMRF %in% c("A1c", "RG", "FG", "PPG") ) %>% arrange(score, CMRF) %>% print(n=40)

p #%>% ggsave(filename="../data/results/descr_plot_cmrf_by_pgs_forgrant.pdf", height=2, width=4)
```

Betas (SE)
```{r}
demo=paste0("age+sex+",paste0("PC",1:10,collapse = "+"))
bmi=paste0(demo, "+bmi")
life=paste0(bmi, "+smoke_level.lab+alch_level.lab+sleep+health")
ses=paste0(life, "+educ_level.lab+townsend")
diet=paste0(ses, "+", paste0(names(diet24hr_vars),collapse = "+"))
covarsets.l <- list(Demographic=demo, BMI=bmi, Lifestyle=life, SES=ses, Diet=diet)

#cmrf_vars <- cmrf_vars[-11]
cmrf_vars <- c(cmrf_vars[-11], glu2hr="2hr-Glucose", bmi="BMI", waist="Waist Circumference")
demo_cmrf_controls <- do.call(rbind.data.frame, lapply(1:length(pgs), function(p) {
  do.call(rbind.data.frame, lapply(1:length(cmrf_vars), function(i) {
    do.call(rbind.data.frame, lapply(1:length(covarsets.l), function(m) {
    print_lm(exposure=pgs[p], outcome=names(cmrf_vars)[i], covariates = covarsets.l[[m]], label = names(covarsets.l[m]), 
             data=analysis %>% filter(t2d_case.f=="Control"))
      }))
    }))
  })) ; demo_cmrf_controls %>% arrange(p)

bitter_a1c <- demo_cmrf_controls %>% arrange(p) %>% filter(outcome=="hba1c" & exposure=="pref_bitter_food_pgs")

bitter_a1c %>% mutate(model=factor(model, levels=names(covarsets.l))) %>%
  ggplot(aes(x=model, y=beta, ymin=beta-se, ymax=beta+se, color=model)) + 
  geom_point() + 
  geom_errorbar() + 
  geom_hline(yintercept = 0)

p <- demo_cmrf_controls %>% filter(model == "Demographic") %>%
  mutate(PGS=factor(exposure, levels=names(pgs_labs), labels=pgs_labs)) %>%
  mutate(dir=ifelse(beta>0,"pos","neg")) %>%
  mutate(CMRF=factor(outcome, levels=names(cmrf_vars), labels=cmrf_vars)) %>%
  ggplot(aes(x=CMRF, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  #facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05), color="black", linetype="longdash") +
  geom_hline(yintercept = -log10(.05/(15)), color="red") +
  geom_point(size=1.55, alpha=0.85) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=as.vector(pgs_palettes)) + 
  scale_color_manual(values=as.vector(pgs_palettes)) + 
   theme(panel.grid.minor =  element_blank(),
        panel.grid.major.x = element_blank()) + 
  coord_flip() ; p

p %>% ggsave(filename="../output/plot_mht_prefPGS_cmrf_testing.pdf", width=6, height=5)


```



#### Preference PGS & Food groups (24HR)
Means/SE
```{r}
analysis_not2d <- analysis %>% filter(t2d_case.f=="Control")
pgs_deciles <- paste0(pgs,"_decile")
diet_vars_cont <- diet_vars[!c(endsWith(names(diet_vars),"lab") | startsWith(names(diet_vars),"pref"))]
diet_pgs_forplot <- do.call(rbind.data.frame, lapply(pgs_deciles, function(pgs_dec) {
  do.call(rbind.data.frame, lapply(1:length(diet_vars_cont), function(d) {
    analysis_not2d %>% rename(dec=pgs_dec) %>% group_by(PGS_decile=all_of(dec)) %>%
      select(PGS_decile, diet=names(diet_vars_cont)[d]) %>%
      reframe(n=n(), mean=mean(diet, na.rm=T), sd=sd(diet, na.rm=T)) %>%
      mutate(se=sd/sqrt(n), 
             Food=diet_vars_cont[d])
  })) %>% mutate(score=pgs_short[gsub("_decile","",pgs_dec)])
}))

## associations
assoc_pgs_diet <- do.call(rbind.data.frame, lapply(1:length(pgs), function(p) {
  do.call(rbind.data.frame, lapply(1:length(diet_vars_cont), function(i) {
    do.call(rbind.data.frame, lapply(1:length(covarsets.l), function(m) {
    print_lm(exposure=pgs[p], outcome=names(diet_vars_cont)[i], covariates = covarsets.l[[m]], label = names(covarsets.l[m]), 
             data=analysis_not2d)
      }))
    }))
  })) ; assoc_pgs_diet %>% arrange(p)
```

```{r}
diet_vars <- diet_vars
## Bitter --> HbA1c
summary(lm(formula(paste0("hba1c~as.numeric(pref_bitter_food_pgs_decile)+",covarsets.l$Demographic)), data=analysis_not2d)) #0.000112 ***
summary(lm(formula(paste0("hba1c~pref_bitter_food_pgs+",covarsets.l$Lifestyle,"+ancestry")), data=analysis_not2d)) #0.000111 ***
summary(lm(formula(paste0("hba1c~pref_bitter_food_pgs+",covarsets.l$Lifestyle,"+ancestry",
                          "+",paste0(names(diet_vars_cont),collapse="+"))), data=analysis_not2d)) #0.000339 ***

## Salt --> CRP
summary(lm(formula(paste0("crp~as.numeric(pref_salty_food_pgs_decile)+",covarsets.l$Demographic)), data=analysis_not2d)) #0.009611 ** 
summary(lm(formula(paste0("crp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry")), data=analysis_not2d)) #0.008309 ** 
summary(lm(formula(paste0("crp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry",
                          "+",paste0(names(diet_vars_cont),collapse="+"))), data=analysis_not2d)) #0.021682 *  


## Salt --> CRP
summary(lm(formula(paste0("dbp~pref_salty_food_pgs+",covarsets.l$Demographic,"+ancestry")), data=analysis_not2d)) #0.009611 ** 
summary(lm(formula(paste0("dbp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry")), data=analysis_not2d)) #0.008309 ** 
summary(lm(formula(paste0("dbp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry",
                          "+",paste0(names(diet_vars_cont),collapse="+"))), data=analysis_not2d)) #0.021682 *  

## Sweet --> HDL
summary(lm(formula(paste0("hdl~as.numeric(pref_sweet_food_pgs_decile)+",covarsets.l$Demographic)), data=analysis_not2d)) #0.009611 ** 
summary(lm(formula(paste0("crp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry")), data=analysis_not2d)) #0.008309 ** 
summary(lm(formula(paste0("crp~pref_salty_food_pgs+",covarsets.l$Lifestyle,"+ancestry",
                          "+",paste0(names(diet_vars_cont),collapse="+"))), data=analysis_not2d)) #0.021682 *  


```
















```{r, fig.width=15, fig.height=6}
pgs_prefs <- fread("../data/processed/results/lm_macroPGS_allPrefs_mBase_testing.csv")
codebook_prefs <- readxl::read_xlsx("../../tasteprefs/data/raw/food_pref_vars.xlsx")

prefs_taste <- codebook_prefs %>% pivot_longer(c(tastes, "nondiet", "spicy")) %>%
  filter(value == 1)

# P-value correction: 153 (-13 nondiet) traits = 140 foods * 3 PGS = 420
 
pgs_prefs_toPlot <- pgs_prefs %>% mutate(PGS=ifelse(macronutrient_pgs=="carb_pgs", names(macro_pgs)[1],
                                ifelse(macronutrient_pgs=="fat_pgs", names(macro_pgs[2]), names(macro_pgs[3])))) %>%
  mutate(dir=ifelse(beta>0,"pos","neg")) %>%
  rowwise() %>%
  mutate(cat = ifelse(outcome %in% prefs_taste$varID, prefs_taste$name[which(prefs_taste$varID==outcome)], "misc")) %>%
  ungroup() %>%
  mutate(cat=factor(str_to_sentence(cat), levels=c("Sweet", "Fatty", "Umami", "Bitter", "Salt", "Sour", "Spicy", "Misc", "Nondiet"))) %>%
  mutate(food=str_to_sentence(gsub("Liking for ", "", model)))

p <- pgs_prefs_toPlot %>%
  ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05/420), color="red") +
  geom_point(size=1.25) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
  scale_color_manual(values=c("#A6501E", "#1D4884","#356932"))

p %>% ggsave(filename="../output/plot_mht_macroPGS_foodprefs_testing.pdf", height=3, width=16)

```


```{r}
#Focus on beta coefficients
pB <- pgs_prefs_toPlot %>%
  ggplot(aes(x=food, y=beta, ymin=beta-se, ymax=beta+se, color=PGS)) + 
  facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") + ylab("Beta (SE)") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_point(size=1.25, position=position_dodge(0.25)) +
  geom_errorbar(width=0.55,  position=position_dodge(0.25)) + 
  scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
  scale_color_manual(values=c("#A6501E", "#1D4884","#356932")) + 
  theme(legend.position = "none")

pB %>% ggsave(filename="../output/plot_dotBeta_macroPGS_foodprefs_testing.pdf", height=3.5, width=16)

```


```{r}
#Zoom in on Nondiet traits
pz <- ggpubr::ggarrange(
  pgs_prefs_toPlot %>%
    filter(cat=="Nondiet") %>%
    ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
    facet_grid(~cat, space = "free", scales = "free_x") +
    theme_bw() + xlab("") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
    geom_hline(yintercept = 0) +
    geom_hline(yintercept = -log10(.05/(13*5)), color="red") +
    geom_point(size=1.25, position=position_dodge(0.35)) + 
    scale_shape_manual(values=c(pos=24, neg=25)) +
    #geom_errorbar() + 
    scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    scale_color_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    theme(legend.position = "none"),
  pgs_prefs_toPlot %>%
    filter(cat=="Nondiet") %>%
    ggplot(aes(x=food, y=beta, ymin=beta-se, ymax=beta+se, color=PGS)) + 
    facet_grid(~cat, space = "free", scales = "free_x") +
    theme_bw() + xlab("") + ylab("Beta (SE)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
    geom_hline(yintercept = 0) +
    geom_point(size=1.25, position=position_dodge(0.35)) +
    ylab("Beta (95% CI)") +
    geom_errorbar(width=0.55,  position=position_dodge(0.35)) + 
    scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    scale_color_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    theme(legend.position = "none"),
  common.legend = T, ncol=2, legend = "none"
)

#pz %>% ggsave(filename="../output/plot_mht_macroPGS_foodprefs_nondiet_testing.pdf", height=3, width=4)

```

#### Addiing adjustment for macronutrient intake

```{r, fig.width=15, fig.height=6}
pgs_prefs_adjMacro <- fread("../data/processed/results/lm_macroPGS_allPrefs_mBase_adjMacro_testing.csv")

prefs_taste <- codebook_prefs %>% pivot_longer(c(tastes, "nondiet", "spicy")) %>%
  filter(value == 1)

pgs_prefs_adj_toPlot <- pgs_prefs_adjMacro %>% mutate(PGS=ifelse(macronutrient_pgs=="carb_pgs", names(macro_pgs)[1],
                                ifelse(macronutrient_pgs=="fat_pgs", names(macro_pgs[2]), names(macro_pgs[3])))) %>%
  mutate(dir=ifelse(beta>0,"pos","neg")) %>%
  rowwise() %>%
  mutate(cat = ifelse(outcome %in% prefs_taste$varID, prefs_taste$name[which(prefs_taste$varID==outcome)], "misc")) %>%
  ungroup() %>%
  mutate(cat=factor(str_to_sentence(cat), levels=c("Sweet", "Fatty", "Umami", "Bitter", "Salt", "Sour", "Spicy", "Misc", "Nondiet"))) %>%
  mutate(food=str_to_sentence(gsub("Liking for ", "", model)))

# p-values
p.adj <- pgs_prefs_adj_toPlot %>%
  ggplot(aes(x=food, y=-log10(p), fill=PGS, color=PGS, shape=dir)) + 
  facet_grid(~cat, space = "free", scales = "free_x") +
  theme_bw() + xlab("") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = -log10(.05/420), color="red") +
  geom_point(size=1.25) + scale_shape_manual(values=c(pos=24, neg=25)) +
  #geom_errorbar() + 
  scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
  scale_color_manual(values=c("#A6501E", "#1D4884","#356932"))

p.adj %>% ggsave(filename="../output/plot_mht_macroPGS_foodprefs_adjMacro_testing.pdf", height=3, width=16)

#betas
p.adjB <- pgs_prefs_adj_toPlot %>%
    ggplot(aes(x=food, y=beta, ymin=beta-se, ymax=beta+se, color=PGS)) + 
    facet_grid(~cat, space = "free", scales = "free_x") +
    theme_bw() + xlab("") + ylab("Beta (SE)") +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=6), strip.background = element_blank()) + 
    geom_hline(yintercept = 0) +
    geom_point(size=1.25, position=position_dodge(0.25)) +
    geom_errorbar(width=0.55,  position=position_dodge(0.25)) + 
    scale_fill_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    scale_color_manual(values=c("#A6501E", "#1D4884","#356932")) + 
    theme(legend.position = "none")

p.adjB %>% ggsave(filename="../output/plot_dotBeta_macroPGS_foodprefs_adjMacro_testing.pdf", height=3.5, width=16)


```







